### 2018-06-19 paper/1-intro.tex
Finish S5

\item Moreover, we can readily remix the ingredients to describe the first
  build system that supports both \emph{dynamic dependencies}
  and \emph{cloud build} (\S\ref{sec-implementation-cloud-shake}).

>>>>

\item Moreover, we can readily remix the ingredients to describe the first
  build system that supports both \emph{dynamic dependencies}
  and \emph{cloud build} (\S\ref{sec-implementation-cloud}).

### 2018-06-19 paper/2-background.tex
Finish S5

In this paper we elucidate which build system properties are consequences of
specific implementation choices (metadata and algorithm), and how one can obtain
new build systems with desired properties by recombining parts of existing
implementations. As a compelling example, we demonstrate how to combine the
advantages of \Shake and \Bazel in a cloud build system with dynamic
dependencies, see~\S\ref{sec-implementation-cloud-shake}.

>>>>

In this paper we elucidate which build system properties are consequences of
specific implementation choices (metadata and algorithm), and how one can obtain
new build systems with desired properties by recombining parts of existing
implementations. As a compelling example, we demonstrate how to combine the
advantages of \Shake and \Bazel in a cloud build system with dynamic
dependencies, see~\S\ref{sec-implementation-cloud}.

### 2018-06-19 paper/4-build.tex
Finish S5

@@

With the information in this section we can build a table comparing the
dependency order strategy with the out-of-date keys strategy, providing 9~possible
build systems, 5~of which are actually inhabited by existing build systems
(we discuss \Ninja \cite{ninja} in \S\ref{sec-related-build}). Of the
remaining 4~spots, we believe neither Traced or Cloud \Excel make sense~--~the
\Excel approach of reordering combined with static approximations reduces
the memory usage significantly. However, as soon as you are paying the cost of
storing traces, that benefit is gone. The advantage of Approximate \Shake over
\Shake would be that it could avoid storing traces and having a separate
information database, but that advantage is minor compared to the technical
restrictions and approximations it would provide, so we consider it unlikely
to be built. The Cloud \Shake system is an interesting and important point in
the design space, which we explore further in~\S\ref{sec-implementation-cloud-shake}.

>>>>

\subsection{Smarter \hs{[Trace]} data structures}\label{sec-smart-traces}

In the examples above, we have used \hs{[Trace}~\hs{k}~\hs{v]} to capture a list
of traces~--~however, using a list necessarily means that finding the right trace
takes $O(n)$. For each of the \hs{Trace} based systems it is possible to devise
a smarter representation, which we sketch below. Note that these implementations
do not avoid calls to \hs{compute}, merely overheads in the build system itself.

\begin{enumerate}
\item Any system using verifying traces, e.g. \Shake, is unlikely to see significant benefit from storing more than one \hs{Trace} per key\footnote{There is a small chance of a benefit if the dependencies change but the result does not, and then the dependencies change back to what they were before.}. Therefore, such systems can store \hs{Map}~\hs{k}~\hs{(Trace}~\hs{k}~\hs{v)}, where the initial \hs{k} is the \hs{key} field of \hs{Trace}.
\item Any system using \hs{Applicative} dependencies can omit the dependency keys from the \hs{Trace} as they can be recovered from the \hs{key} field.
\item Any \hs{Applicative} build system storing constructive traces, e.g. \Bazel, can index directly from the key and results to the output result~--~i.e. \hs{Map}~\hs{(@@k,}~\hs{[Hash}~\hs{v])}~\hs{(Hash}~\hs{v)}. Importantly, assuming the traces are stored on a central server, the client can compute the key and the hashes of its dependencies, then make a single call to the server to retrieve the result hash. In this formulation we have removed the possibility for a single key/dependency state to map to multiple different hashes, e.g. on a non-deterministic build~--~something \Bazel already prohibits which is discussed more in~\S\ref{sec-non-determinism}.
\item Finally, a \hs{Monad} build system with constructive traces can be stored as \hs{Map}~\hs{k}~\hs{(Choice}~\hs{k}~\hs{v)}, assuming a definition of \hs{Choice} as:
\begin{minted}[xleftmargin=10pt]{haskell}
data Choice k v = Choice k (Map (Hash v) (Choice k v))
                | Result (Hash v)
\end{minted}
Here the \hs{Choice} encodes a tree, asking successive questions about keys, and taking different branches based on the answers, until it reaches a final result. Implementing this structure over client-server communication requires either a chatty interface with lots of round-trips per \hs{Choice} step, or sending over a part of the tree that is not subsequently explored.
\end{enumerate}

@@

With the information in this section we can build a table comparing the
dependency order strategy with the out-of-date keys strategy, providing 9~possible
build systems, 5~of which are actually inhabited by existing build systems
(we discuss \Ninja \cite{ninja} in \S\ref{sec-related-build}). Of the
remaining 4~spots, we believe neither Traced or Cloud \Excel make sense~--~the
\Excel approach of reordering combined with static approximations reduces
the memory usage significantly. However, as soon as you are paying the cost of
storing traces, that benefit is gone. The advantage of Approximate \Shake over
\Shake would be that it could avoid storing traces and having a separate
information database, but that advantage is minor compared to the technical
restrictions and approximations it would provide, so we consider it unlikely
to be built. The Cloud \Shake system is an interesting and important point in
the design space, which we explore further
in~\S\ref{sec-implementation-cloud}.

### 2018-06-19 paper/5-implementations.tex
Finish S5

\subsection{Cloud build systems: \Bazel and \Cloud \Shake}\label{sec-implementation-bazel-cloud-shake}

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Bazel build system, implementation of 'restarting2' is omitted (22 lines)
bazel :: (Ord k, Hashable v) => Build Monad (CT k v) k v
bazel = restarting2 ctRebuilder
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Cloud Shake build system, implementation of 'suspending' is given in Fig. 8
cloudShake :: (Ord k, Hashable v) => Build Monad (CT k v) k v
cloudShake = suspending ctRebuilder
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A rebuilder based on constructive traces
ctRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (CT k v) k v
ctRebuilder key value task = Task $ \fetch -> do
    maybeCachedValue <- constructCT key value (fmap hash . fetch) =<< get
    case maybeCachedValue of
        Just cachedValue -> return cachedValue
        Nothing -> do
            (newValue, deps) <- track task fetch
            put =<< recordCT key newValue deps (fmap hash . fetch) =<< get
            return newValue
\end{minted}
\vspace{-2mm}
\caption{Implementations of \Bazel and \Cloud \Shake using our framework;
\hs{restarting2} differs from the restarting scheduler used by Excel
(Fig.~\ref{fig-excel-implementation}) in that it does not rely on the
calculation chain.}
% \hs{topological} is defined in Fig.~\ref{fig-make-implementation}.
\label{fig-bazel-cloud-shake-implementation}
\vspace{-2mm}
\end{figure}

In this section we use constructive traces to implement cloud build systems.
Fig.~\ref{fig-bazel-cloud-shake-implementation} shows our models of \Bazel and
\Cloud \Shake, which both rely on the \hs{ctRebuilder} -- a rebuilder based on
constructive traces (\S\ref{sec-constructive-traces}).

@@

Note that although both our model and \Bazel's actual implementation supports
dynamic dependencies, it is currently not possible to define new monadic build
rules in the language available to users. Instead, users have to rely on a
(rich) collection of predefined built-in rules, which cover most but not all
practically important cases of dynamic dependencies.

@@

\textbf{TODO:} A real implementation of Cloud \Shake is now available... It was
obtained by directly applying the presented ideas.

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Buck build system, implementation of 'topological' is given in Fig. 6
buck :: (Hashable k, Hashable v) => Build Applicative (DCT k v) k v
buck = topological (unliftRebuilder dctRebuilder)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Convert a monadic rebuilder to the corresponding applicative one
unliftRebuilder :: Rebuilder Monad i k v -> Rebuilder Applicative i k v
unliftRebuilder rebuilder key value task = rebuilder key value $ Task $ run task
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Nix build system, implementation of 'suspending' is given in Fig. 8
nix :: (Hashable k, Hashable v) => Build Monad (DCT k v) k v
nix = suspending dctRebuilder
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Rebuilder based on deterministic constructive traces
dctRebuilder :: (Hashable k, Hashable v) => Rebuilder Monad (DCT k v) k v
dctRebuilder key _value task = Task $ \fetch -> do
    dct <- get
    maybeCachedValue <- constructDCT key (fmap hash . fetch) dct
    case maybeCachedValue of
        Just cachedValue -> return cachedValue
        Nothing -> do
            (newValue, deps) <- trackM task fetch
            put =<< recordDCT key newValue deps (fmap hash . fetch) =<< get
            return newValue
\end{minted}
\vspace{-2mm}
\caption{Implementations of \Buck and \Nix using our framework.}
\label{fig-buck-nix-implementation}
\vspace{-2mm}
\end{figure}

@@

The program presented above captures certain aspects of \Bazel, but the real
implementation makes one important additional assumption on \hs{Task} -- namely
that it is \textit{deterministic}. With that assumption \Bazel is able to create
the result hash in a trace by hashing together the result hashes of the
dependencies and the key -- as the mapping between dependencies and results is
deterministic. As a consequence \Bazel can compute the results of traces locally,
without looking at \hs{Traces} (potentially saving a roundtrip to the server).
To model this change in the code would require storing an additional transient
piece of information \hs{done} of type \hs{Map}~\hs{k}~\hs{(Hash}~\hs{v)},
storing the computed hashes so far this run. With that available,
\hs{getHash}~\hs{key}~\hs{s} would become:
\begin{minted}[xleftmargin=10pt]{haskell}
hash (key, [ done Map.! d | d <- ds ])
\end{minted}
And that new hash would have to be stored in \hs{done}.

\vspace{-1mm}
\subsection{Smarter \hs{[Trace]} data structures}\label{sec-smart-traces}
\vspace{-1mm}

In the examples above, we have used \hs{[Trace}~\hs{k}~\hs{v]} to capture a list
of traces~--~however, using a list necessarily means that finding the right trace
takes $O(n)$. For each of the \hs{Trace} based systems it is possible to devise
a smarter representation, which we sketch below. Note that these implementations
do not avoid calls to \hs{compute}, merely overheads in the build system itself.

\begin{enumerate}
\item Any system using verifying traces, e.g. \Shake, is unlikely to see significant benefit from storing more than one \hs{Trace} per key\footnote{There is a small chance of a benefit if the dependencies change but the result does not, and then the dependencies change back to what they were before.}. Therefore, such systems can store \hs{Map}~\hs{k}~\hs{(Trace}~\hs{k}~\hs{v)}, where the initial \hs{k} is the \hs{key} field of \hs{Trace}.
\item Any system using \hs{Applicative} dependencies can omit the dependency keys from the \hs{Trace} as they can be recovered from the \hs{key} field.
\item Any \hs{Applicative} build system storing constructive traces, e.g. \Bazel, can index directly from the key and results to the output result~--~i.e. \hs{Map}~\hs{(@@k,}~\hs{[Hash}~\hs{v])}~\hs{(Hash}~\hs{v)}. Importantly, assuming the traces are stored on a central server, the client can compute the key and the hashes of its dependencies, then make a single call to the server to retrieve the result hash. In this formulation we have removed the possibility for a single key/dependency state to map to multiple different hashes, e.g. on a non-deterministic build~--~something \Bazel already prohibits which is discussed more in~\S\ref{sec-non-determinism}.
\item Finally, a \hs{Monad} build system with constructive traces can be stored as \hs{Map}~\hs{k}~\hs{(Choice}~\hs{k}~\hs{v)}, assuming a definition of \hs{Choice} as:
\begin{minted}[xleftmargin=10pt]{haskell}
data Choice k v = Choice k (Map (Hash v) (Choice k v))
                | Result (Hash v)
\end{minted}
Here the \hs{Choice} encodes a tree, asking successive questions about keys, and taking different branches based on the answers, until it reaches a final result. Implementing this structure over client-server communication requires either a chatty interface with lots of round-trips per \hs{Choice} step, or sending over a part of the tree that is not subsequently explored.
\end{enumerate}

>>>>

\subsection{Cloud build systems: \Bazel, \CloudBuild and \Cloud \Shake}\label{sec-implementation-cloud}

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Bazel build system, implementation of 'restarting2' is omitted (22 lines)
bazel :: (Ord k, Hashable v) => Build Monad (CT k v) k v
bazel = restarting2 ctRebuilder
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A rebuilder based on constructive traces
ctRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (CT k v) k v
ctRebuilder key value task = Task $ \fetch -> do
    maybeCachedValue <- constructCT key value (fmap hash . fetch) =<< get
    case maybeCachedValue of
        Just cachedValue -> return cachedValue
        Nothing -> do
            (newValue, deps) <- track task fetch
            put =<< recordCT key newValue deps (fmap hash . fetch) =<< get
            return newValue
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- CloudBuild build system, implementation of 'topological' is given in Fig. 6
cloudBuild :: (Ord k, Hashable v) => Build Applicative (CT k v) k v
cloudBuild = topological (adaptRebuilder ctRebuilder)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Convert a monadic rebuilder to the corresponding applicative one
adaptRebuilder :: Rebuilder Monad i k v -> Rebuilder Applicative i k v
adaptRebuilder rebuilder key value task = rebuilder key value $ Task $ run task
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Cloud Shake build system, implementation of 'suspending' is given in Fig. 8
cloudShake :: (Ord k, Hashable v) => Build Monad (CT k v) k v
cloudShake = suspending ctRebuilder
\end{minted}
\vspace{-2mm}
\caption{Implementations of \Bazel, \CloudBuild and \Cloud \Shake using our
framework; \hs{restarting2} differs from the restarting scheduler used by \Excel
(Fig.~\ref{fig-excel-implementation}) in that it does not rely on the
calculation chain.}
\label{fig-bazel-cloud-shake-implementation}
\vspace{-4mm}
\end{figure}

In this section we use constructive traces to implement cloud build systems.
Fig.~\ref{fig-bazel-cloud-shake-implementation} shows our models of \Bazel,
\CloudBuild and \Cloud \Shake, which both rely on the \hs{ctRebuilder} -- a
rebuilder based on constructive traces (\S\ref{sec-constructive-traces}).

@@

Note that although both our model and \Bazel's actual implementation supports
dynamic dependencies, it is currently not possible to define new monadic build
rules in the language available to users. Instead, users have to rely on a
(rich) collection of predefined built-in rules, which cover many but not all
practically important cases of dynamic dependencies.

By switching to the \hs{topological} scheduler, we obtain a model of
Microsoft's \CloudBuild~-- an applicative build system that combines
conventional scheduling of statically known directed acyclic graphs of
dependencies with constructive traces~\cite{esfahani2016cloudbuild}. Note that
we need to convert a monadic \hs{ctRebuilder} into an applicative one by
applying an adapter function \hs{adaptRebuilder}, which unwraps a given
\hs{Task Applicative} and wraps it into \hs{Task Monad}.

@@

\textbf{TODO:} A real implementation of \Cloud \Shake is now available... It was
obtained by directly applying the presented ideas.

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Buck build system, implementation of 'topological' is given in Fig. 6
buck :: (Hashable k, Hashable v) => Build Applicative (DCT k v) k v
buck = topological (adaptRebuilder dctRebuilder)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Nix build system, implementation of 'suspending' is given in Fig. 8
nix :: (Hashable k, Hashable v) => Build Monad (DCT k v) k v
nix = suspending dctRebuilder
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Rebuilder based on deterministic constructive traces
dctRebuilder :: (Hashable k, Hashable v) => Rebuilder Monad (DCT k v) k v
dctRebuilder key _value task = Task $ \fetch -> do
    dct <- get
    maybeCachedValue <- constructDCT key (fmap hash . fetch) dct
    case maybeCachedValue of
        Just cachedValue -> return cachedValue
        Nothing -> do
            (newValue, deps) <- track task fetch
            put =<< recordDCT key newValue deps (fmap hash . fetch) =<< get
            return newValue
\end{minted}
\vspace{-2mm}
\caption{Implementations of \Buck and \Nix using our framework.}
\label{fig-buck-nix-implementation}
\vspace{-2mm}
\end{figure}

@@

Some build systems allow to exploit \emph{build task determinism} for efficient
fetching of build results from the cache: one can simply hash the values of all
transitive inputs of a key (which can be done efficiently by storing all hashes
in a Merkle tree) and download the matching value from the cache, without the
need to build any intermediate values. This can bring significant performance
benefits in practice, but sacrifices the early cutoff property: indeed, with
this approach addition of a new comment to one of the source files will
invalidate all dependent keys.

Fig.~\ref{fig-buck-nix-implementation} shows our models of \Buck and \Nix build
systems that use the \hs{dctRebuilder} -- a rebuilder based on deterministic
constructive traces. As one can see, \hs{dctRebuilder} is almost identical to
\hs{ctRebuilder}, but it ignores the current \hs{value}, since the
\hs{constructDCT} query only needs to examine the transitive inputs of the given
\hs{key}. \Buck uses the \hs{topological} scheduler and is therefore an
applicative build system, whereas \Nix uses the \hs{suspending} scheduler and is
monadic.

### 2018-06-19 paper/7-related.tex
Finish S5

\item \CloudBuild~\cite{esfahani2016cloudbuild}
differs from \Bazel by allowing non-determinism~\S\ref{sec-non-determinism}, thus more closely modelling
our original definition of \Bazel from \S\ref{sec-implementation-bazel-cloud-shake}.

\item \Nix~\cite{dolstra2004nix} has coarse-grained dependencies, with precise
hashing of dependencies and downloading of precomputed build products. When
combined with \cmd{import-from-derivation}, \Nix can also be considered monadic,
making it similar to \Cloud \Shake
from~\S\ref{sec-implementation-bazel-cloud-shake}.
However, \Nix is not intended as a build system, and the coarse grained nature
(packages, not individual files) makes it targeted to a different purpose.
% John Ericson suggested in a blog comment that nix may be somewhat monadic, see:
% \url{https://blogs.ncl.ac.uk/andreymokhov/cloud-and-dynamic-builds/\#comment-1849}.

>>>>

\item \CloudBuild~\cite{esfahani2016cloudbuild}
differs from \Bazel by allowing non-determinism~\S\ref{sec-non-determinism}, thus more closely modelling
our original definition of \Bazel from \S\ref{sec-implementation-cloud}.

\item \Nix~\cite{dolstra2004nix} has coarse-grained dependencies, with precise
hashing of dependencies and downloading of precomputed build products. When
combined with \cmd{import-from-derivation}, \Nix can also be considered monadic,
making it similar to \Cloud \Shake
from~\S\ref{sec-implementation-cloud}.
However, \Nix is not intended as a build system, and the coarse grained nature
(packages, not individual files) makes it targeted to a different purpose.
% John Ericson suggested in a blog comment that nix may be somewhat monadic, see:
% \url{https://blogs.ncl.ac.uk/andreymokhov/cloud-and-dynamic-builds/\#comment-1849}.
